---
title: "Estimation of Transition Matrices"
subtitle: "Based on @Craig2002"
author: "Fernando Alarid-Escudero & Karen M Kuntz"
date: "February 12, 2015"
output: beamer_presentation
references:
- id: Craig2002
  title: Estimation of the transition matrix of a discrete-time Markov chain
  author:
    - family: Craig
      given: Bruce A 
    - family: Sendi
      given: Peter P
  container-title: Health Economics
  volume: 42
  issue: 11
  page: 33-42
  type: article-journal
  issued:
    year: 2002
- id: Cinlar1975
  title: Introduction to Stochastic Processes
  author:
    - family: Cinlar
      given: Erhan  
  publisher: Prentice-Hall, Inc
  page: 364-367
  type: book
  issued:
    year: 1975    
---

## Transition probability matrix
A Markov model with $M$ states will have the transition matrix
$$
P=
\begin{bmatrix}
p_{1,1} & p_{1,2} & \cdots & p_{1,M} \\
p_{2,1} & p_{2,2} & \cdots & p_{2,M} \\
\vdots & \vdots & \ddots & \vdots \\
p_{M,1} & p_{M,2} & \cdots & p_{M,M}
\end{bmatrix},
$$
where $p_{i,j}$ is the trasnistion probability from state $i$ to state $j$ in period $T$.

## Transition count matrix
The transition count matrix represents the observed transitions in period $T$ and is expressed as
$$
N=
\begin{bmatrix}
n_{1,1} & n_{1,2} & \cdots & n_{1,M} \\
n_{2,1} & n_{2,2} & \cdots & n_{2,M} \\
\vdots & \vdots & \ddots & \vdots \\
n_{M,1} & n_{M,2} & \cdots & n_{M,M}
\end{bmatrix},
$$
where $n_{i,j}$ is the number of occurrences between state $i$ and state $j$ in period $T$.

## Estimation of $P$ from longitudinal cohort data
The maximum likelihood estimation (MLE) of $P$ is simply the row proportions of $N$
$$
\hat{P}=\{\hat{\textbf{p}}\} \text{ where } \hat{p}_{i,j}=\frac{n_{i,j}}{\sum_{m=1}^{M}{n_{i,m}}}
$$

## Estimate of $P$ for a different time scale 
- If the observation periods of the longitudinal cohort study are carried out every $t_0$, then the MLE estimate of its corresponding transition probability matrix is $\hat{P}_{t_0}$. 
- You are interested in a transition matrix for a different time scale (e.g., monthly or biannual)
- If $t_0$ is the original time scale and $t_f$ the desired time scale, by the invariance property, the MLE of $P_{t_f}$ associated with a cycle length of $t_f$ is simply 
$$
\hat{P}_{t_f} = \hat{P}_{t_0}^{(t_f/t_0)}
$$
- For example, to obtain a one-year transition matrix from a one-month transition matrix we just simply raise the one-month transition matrix to the twelfth power
$$
\hat{P}_{yr} = \hat{P}_{mo}^{(12)}
$$

## Potential issues 
- It is relatively easy to go from a shorter time scale to a longer time scale if the latter is a multiple of the former (i.e., $(t_f/t_0)$ is a positive integer)
- In our monthly to annual scale example we just simply multiply $\hat{P}_{mo}$ 12 times!
$$
\hat{P}_{yr} = \hat{P}_{mo} \cdots \hat{P}_{mo}
$$
- If this is not the case, we need to do eigenvalues-eigenvectors (spectral) decomposition.

## Spectral decomposition
- If $P$ is our $M \times M$ transition probability matrix, then its spectral decomposition is defined by
$$
P = V D V^{-1}, \text{ where } D = 
\begin{bmatrix}
\lambda_{1} & 0 & \cdots & 0 \\
0 & \lambda_{2} & \ddots & \vdots \\
\vdots & \ddots & \ddots & 0 \\
0 & \cdots & 0 & \lambda_{M}
\end{bmatrix},
$$
and $\lambda_i$ is the $i$th eigenvalue and its associated eigenvector is the $i$th column of $V$.

## Spectral decomposition [@Cinlar1975]
- If we want a transition matrix $P_t$ in a time scale that is $t$ times the original scale (where $t$ could be an integer or not ), then it follows that
$$
P_t = V D^t V^{-1}, \text{ where } D^{t} = 
\begin{bmatrix}
\lambda^{t}_{1} & 0 & \cdots & 0 \\
0 & \lambda^{t}_{2} & \ddots & \vdots \\
\vdots & \ddots & \ddots & 0 \\
0 & \cdots & 0 & \lambda^{t}_{M}
\end{bmatrix},
$$
- The eigenvalues are raised to the power $t$ but the eigenvectors do not change.

## Example in `R`
Install `markovchain` package to obtain example data
```{r, eval=FALSE}
install.packages{markovchain}
```
Load package and example data `craigsendi` from @Craig2002. These data comes from the Swiss HIV cohort study and describes the progression of HIV-infected subjects in terms of their observed six-month CD4-cell count transitions
```{r, message=FALSE}
require(markovchain)
data(craigsendi)
N<-craigsendi
```

## Example

Display the 6-month transition count matrix $N$
```{r}
N
```

## Example

Initialize a $3\times 3$ matrix
```{r}
P <- matrix(nrow = 3, ncol = 3, 0)
colnames(P)<-rownames(P)<-colnames(N)
P
```

## Example

Compute the MLE of the 6-month transition probability matrix $P$ from transition count matrix $N$. 
```{r}
for (i in 1:3) P[i, ] <- N[i, ] / sum(N[i, ])
P
```

## Example

Compute one-month transition probability matrix $P_1$ from the 6-month transtion matrix $P$ using the eigen vectors and eigen values decomposition
```{r}
eig <- eigen(P)
eig
```

## Example
- Create diagonal matrix D with eigen values
```{r}
D <- diag(eig$values)
```

- Create matrix $V$ with eigen vectors
```{r}
V <- eig$vectors
```

## Example

- Show that both matrices can recreate $P$
```{r}
V %*% D %*% solve(V)
```
and they do!

- Take the sixth root of each of the diagonal's element of matrix D
```{r}
d <- D^(1/6)
```

## Example

Generate the one-month transition matrix
```{r}
P1 <- V %*% d %*% solve(V)
P1
```

## References

